Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2016-07-25 document/ms.tex
Some edits through the results sections, trim some extraneous paragraphs

\newcommand{\iancze}{{\sc C15 }}

@@

Starspots have recently been observed with exceptional photometric precision from monitoring surveys targeting exoplanet transits \citep{harrison11,davenport15}.  Contemporaneous or near-contemporaneous panchromatic photometric monitoring \citep{bouvier95,grankin07} can break the degeneracy between starspot areal filling factor and starspot temperature. Photometric measurements are limited to detecting the largest longitudinally asymmetric starspots.  Transiting planets or eclipsing binaries passing in front of starspots \citep{desert11} can break geometric degeneracies, providing estimates for the characteristic sizes and lifetimes of starspots or starspot groups.  Zeeman Doppler Imaging (ZDI, \emph{e.g.} \citet{donati14}) suffers from the same longitudinal asymmetry limitation as photometry, but can yield information on the latitudinal distribution of starspots.  All of these techniques would underestimate the starspot areal coverage of pole-on stars, stars with longitudinally symmetric bands of starspots (e.g. bands), or stars with an isotropic distribution of small starspots.

Spectroscopic strategies have the power to measure both the starspot areal coverage and temperature in any geometric configuration.  At low resolution, SED modeling \citep{wolk96} has detected evidence for sunspots in weak-lined T-Tauri stars (WTTS).  At high resolution, line depth ratio analysis has historically been demonstrated in relatively small bandwidths, namely the TiO bands \citep{neff95,oneal96,oneal98,oneal04}, or the OH 1.563 $\mu$m feature \citep{oneal01}.

@@

In this work we constrain the starspot properties of \name with 4 complementary techniques.  Section 2 and Appendix YY describe a spectral inference technique aimed at generalizing line depth ratio analysis.  Section 3 applies the spectral inference technique to a high resolution panchromatic near-IR spectrum of \name.  Sections 4.1, 4.2, and 4.3 show results from polychromatic photometric monitoring, SED fitting, and optical TiO line-depth ratio fitting.  All lines of evidence are ultimately combined to build a consistent picture of the spectral and temporal evolution of \name, and what can be understood about pre-MS stellar evolution from this exemplar.

@@

\iancze developed a modular framework to infer stellar properties from high resolution spectra.  In this section, we describe how we extended this framework to include spots in a 2-temperature fit to spectra.  Further details of this methodology are described in Appendix \ref{methods-details}.

We experimented with several extensions to the modular framework for spectral inference described in \iancze \footnote{The open source codebase and its full revision history is available at https://github.com/iancze/Starfish.  The experimental fork discussed in this paper is at https://github.com/gully/Starfish}.  The \iancze technique forward models ``emulated'' synthetic spectra from pre-computed model grids, taking into account the discretization error attributable to the coarsely sampled stellar intrinsic parameters.  We use the \PHOENIX~ grid of pre-computed synthetic stellar spectra that span a wide range of wavelengths at high spectral resolution with sampling of 100 K in $\teff$ in our range of interest \citep{husser13}.  \todo{Put in a hyperlink?} We made two principal changes to the modular framework: 1) the introduction of a starspot mixture model characterized by a second $\teff$ and areal fill factor $f$, and 2) the switch from Gibbs sampling to ensemble sampling with chunking by spectral order.  We discuss these two changes below.

We assume the stellar photosphere is characterized as two photospheric components with different effective temperatures, $\teffa$ and $\teffb$, and scalar solid angular coverages $\Omega_a$ and $\Omega_b$, but sharing all the same intrinisic and extrinsic stellar parameters otherwise.  Following and extending the notation of \iancze, we can write down the mixture model as:

@@

The observed filling factor of photosphere $c$ is simply the ratio:

@@

The second important change to the spectral inference framework involves technical aspects of switching from sampling the nuisance and stellar parameters separately in a blocked Gibbs framework to ensemble sampling using \texttt{emcee} \citep{foreman13}.  The affine-invariant \texttt{emcee} ensemble sampler is more resilient to correlations expected among stellar temperatures and fill factor than the Metropolis-Hastings sampler used in Gibbs sampling.  The practical effect of this switch is that we obtain unique stellar parameter estimates $\vT_{o}$ for each spectral order $o$, whereas the \iancze strategy had the power to provide a single set of stellar parameters $\vT$ that was based on all $N_{\rm ord}$ spectral orders.  We combine the $N_{\rm ord}$ sets of inferences on $\vT_{o}$ with heuristics like taking the weighted average of point estimates of parameter distributions of the most reliable subsets of orders.  We assume that these lossy heuristics offer a regression to the mean with a large number of spectral orders.

@@

\subsection{IGRINS Spectroscopy}\label{sec:igrins} 
We acquired observations with the Immersion Grating Infrared Spectrograph, IGRINS \citep{park14,gully12} on the Harlan J. Smith Telescope at McDonald Observatory on 2015-11-18 $09^h$ UTC.  IGRINS is a high resolution near-infrared echelle spectrograph providing simultaneous $R\simeq45,000$ spectra over 1.48-2.48\um.  The spectrograph has two arms with 28 orders in $H-$band and 25 orders in $K-band$.  The data were reduced with the Pipeline Package\footnote{\url{https://github.com/igrins/plp}}.  We performed telluric correction by dividing the spectrum of \name by an A0V star observed near-contemporaneously at similar air-mass.  The broad hydrogen lines in the A0V star produced broad flux excesses in the spectrum of \name.  No effort was made to remove these flux excesses, which affect several spectral orders. 
%List the spectral orders these lines affect.

\subsection{ESPaDOnS Spectroscopy}
We used ESPaDOnS on CFHT to obtain twelve high resolution optical spectra of \name from 8-21 Jan.~2014 as part of the MATYSSE Large Program.  These spectra cover 3900-10000 \AA\ at $R\sim68,000$ and were obtained in spectropolarimetry mode.  The Zeeman Doppler Imaging obtained from these observations were analyzed by \citet{donati14}.  In this paper we concentrate on the intensity spectrum.

Chunking, estimate flux calibration...

@@

We performed spectral fitting on the ESPaDoNs spectrum acquired on 2014 January 11, the fourth spectrum from Table \ref{tbl_estimated_V}.  We chunked the spectrum into a subset of $N_{ord}=$26 spectral orders and performed full-spectrum fitting separately on each of these 26 spectral orders following the Metropolis-Hasting MCMC sampling procedure described in \iancze.  We employed the \PHOENIX model grid with a search range of $\logg \in [3.5, 4.0]$, $\Z \in [-0.5, 0.5]$, and $\teff \in [3500, 4200]$ and trained a spectral emulator as described in \iancze.

We tuned the Metropolis-Hasting step sizes with several iterations of burn-in procedure and visually checked the final chains for convergence.  Our number of samples was generally much longer than the estimated integrated autocorrelation length.  We spot-checked full-spectrum fits in comparison with the observed spectrum, finding modest agreement with the exception of spectral line outliers.  No attempt was made to downweight spectral-line outliers as described in \iancze.  We attempted fits to 9 more of the 35 spectral orders available for this ESPaDoNs spectrum, but the MCMC computations for these orders failed to converge, possibly due to numerical artifacts arising from poor model fits.

We computed the median value and 5$^{th}$ and $95^{th}$ percentiles of burned-in subsets of the MCMC samples described above.  Overall these 26 sets of point estimates for $\vT_{m}$, $\vP_m$ show relatively good agreement, with exceptions.  We have little constraint on $\logg$ and $\Z$ since full-spectrum fitting tends to wash out the weak signal contained in line wings for spectra with a finite level of random noise and systematic error.  Meanwhile, strong metal lines can bias estimates of $\Z$ at the expense of estimates for $\teff$.  We get consistent estimates for\footnote{Note, $v_z$ is an uncorrected recessional velocity, not a proper radial velocity} $v_z \sim 15.6 \pm 0.6$ km/s and $\vsini \sim 28.0 \pm 0.9$ km/s, with the exception of two spectral orders at $6920 <\lambda \;(\mathrm{nm})< 7057$ and $8473 < \lambda \;(\mathrm{nm}) < 8706$.  These two outlying spectral orders converged to unreasonably large values for $\vsini$ and $a_{\rm G}$, the amplitude of the Gaussian process covariance kernel.  One of these orders has an unusually small value for effective temperature, $\teff\sim3500$, comparable to the lower limit of the employed model grid.  Visual comparison of the observed spectrum for a range of fits for these orders shows that no single emulated spectrum could reasonably fit their spectrum because there are numerous large spectral outliers.  The more-well-behaved spectral orders show effective temperature point estimates in the range $\teff=4000\pm130$ K.  Figure \ref{fig:SingleTeffvsOrder} displays the $\teff$ point estimates with 5$^{th}$ and $95^{th}$ percentile error bars placed at the central wavelength of each of the 26 spectral orders for the subset of ESPaDoNs orders that received full-spectrum fitting.

@@

The effective temperature derived at different wavelength intervals exhibit a $\Delta \teff \sim 800$ K spread.  To put this level of effective temperature in context, authors using 4200 K and 3300 K values would conclude evolutionary model derived masses of XX or YY as shown in Figure XX. 
\todo{Put in the numbers and HR diagram figure here.}

It is clear that the single-temperature model is a poor fit to the data.  Our model assumptions were incorrect: no single temperature can describe all the spectral lines present in the spectra.  This discrepancy is circumstantial evidence for the direct detection of spectral lines attributable to starspots.  This circumstantial evidence should not be surprising given the previous evidence that \name possesses a large coverage fraction of starspots \citep{donati14}, along with its large amplitude of photometric variations, as shown in Figure \ref{fig:PhotPhase} and discussed previously by \citet{grankin08}.

We performed a further plausibility check to constrain the effect size of starspots.  We generated flux calibrated spectra at two temperatures, and forward-modeled them to resemble the LkCa4 IGRINS and ESPaDOnS spectra in all other ways (\emph{i.e.} $\logg, \Z, \vt_{\rm ext}$), including noise and calibration parameters.  We coadded the spectra in a mixture model:  $ \mathsf{M}_{mix} = c \cdot \mathsf{M}_A(T_\textrm{eff,A}) + (1-c) \cdot \mathsf{M}_B(T_\textrm{eff,B})$.  We chose a fill factor of starspots of 30\%.  We then re-ran our single-temperature fitting procedure on the two-temperature, synthetic, noised-up data to see what stellar parameters our full-spectrum fitting procedure would na\"{\i}vely derive.  
\comm{I don't think I ever put together a figure for this, should we just skip it?  I think so.}

@@

Some care should be taken when directly comparing results between the ESPaDoNs and IGRINS spectra since they were not taken at the same time.  These spectra have comparable albeit different $\hat V$ of 12.90 and 12.83 for the ESPaDoNs and IGRINS spectra respectively.  If we assume the areal coverage fraction of spots was the same or greater during the ESPaDoNs spectrum acquisition, we can derive the result that the optical bands and short-wavelength end of $H$-band are relatively insensitive to the spectral signatures of starspots in this system.  The long wavelength portion of $H-$band and all of $K-$band are more sensitive to starspot spectral signatures than the shorter wavelength portions.

We can understand the heightened sensitivity to starspot spectral lines as wavelength increases in the following way.  The effective temperature of a starspot is, by definition, cooler than its surrounding photosphere, and will therefore have a longer wavelength of peak emission, according to Wein's displacement law.  So the ratio of flux density between starspot and bulk photosphere as a function of wavelength will generally increase with wavelength until it assymptotes to a fixed value in the Rayleigh-Jeans tail.  In Figure XX we plot the ratio of black body flux density at temperatures 4100 K and 3300 K with a red line.  The blue line is the ratio of two smoothed \PHOENIX model spectra with 4100 K and 3300 K, solar metalicity, and $\logg=3.5$.  The flux density ratio is the flux per solid angle, so the delivered flux ratio for a finite sized starspot will depend on the (flux-weighted) solid angle of the starspot relative to the (flux-weighted) solid angle of the stellar disk, vis-a-vis Equations \ref{eqn:Mix_M} and \ref{eqn:fill_factor}.

\todo{Put in the flux density ratio plot.}

>>>>

\newcommand{\iancze}{{\sc C15}}

@@

Starspots have recently been observed with exceptional photometric precision from monitoring surveys targeting exoplanet transits \citep{harrison11,davenport15}.  Contemporaneous or near-contemporaneous panchromatic photometric monitoring \citep{bouvier95,grankin07} can break the degeneracy between starspot areal filling factor and starspot temperature. Photometric measurements are limited to detecting large longitudinally asymmetric starspots.  Transiting planets or eclipsing binaries passing in front of starspots \citep{desert11} can break geometric degeneracies, providing estimates for the characteristic sizes and lifetimes of starspots or starspot groups.  Zeeman Doppler Imaging \citep[ZDI]{donati14} suffers from the same longitudinal asymmetry limitation as photometry, but can yield information on the latitudinal distribution of starspots.  All of these techniques would underestimate the starspot areal coverage of pole-on stars, stars with longitudinally symmetric bands of starspots, or stars with an isotropic distribution of small starspots.

Spectroscopic strategies have the power to measure both the starspot areal coverage and temperature in any geometric configuration.  At low resolution, SED modeling \citep{wolk96} has detected evidence for sunspots in weak-lined T-Tauri stars (WTTS).  At high resolution, line depth ratio analysis has historically been demonstrated on evolved stars in relatively small bandwidths, namely the TiO bands \citep{neff95,oneal96,oneal98,oneal04}, or the OH 1.563 $\mu$m feature \citep{oneal01}.

@@

In this work we constrain the starspot properties of \name with 4 complementary techniques.  Section 2 and Appendix \ref{methods-details} describe a spectral inference technique aimed at generalizing line depth ratio analysis.  Section 3 applies the spectral inference technique to a high resolution panchromatic near-IR spectrum of \name.  Sections 4.1, 4.2, and 4.3 show results from polychromatic photometric monitoring, SED fitting, and optical TiO line-depth ratio fitting.  All lines of evidence are ultimately combined to build a consistent picture of the spectral and temporal evolution of \name, and what can be understood about pre-MS stellar evolution from this exemplar.

@@

\citet[hereafter \iancze]{czekala15} developed a modular framework\footnote{The open source codebase and its full revision history is available at \url{https://github.com/iancze/Starfish}.  The experimental fork discussed in this paper is at \url{https://github.com/gully/Starfish}} to infer stellar properties from high resolution spectra.  The \iancze technique forward models ``emulated'' synthetic spectra from pre-computed model grids, taking into account the discretization error attributable to the coarsely sampled stellar intrinsic parameters.  We employed the \PHOENIX~ grid of pre-computed synthetic stellar spectra that span a wide range of wavelengths at high spectral resolution with sampling of 100 K in $\teff$ in our range of interest \citep{husser13}.  The modular framework was altered to accommodate starspot measurements in two ways. First, the single photospheric component was updated to includes a starspot spectrum. Second, the MCMC sampling strategy was altered to accommodate the additional free parameters added in the starspot model.

The stellar photosphere is characterized as two photospheric components with a starspot temperature $\teffb$ and ambient photosphere temperature $\teffa$, with scalar solid angular coverages $\Omega_b$ and $\Omega_a$, respectively.    Starspots (or spot groups) are assumed to be cooler than the ambient photosphere, and they share the same average intrinisic and extrinsic stellar parameters otherwise; that is $(\vsini, \logg, \Z, v_z)$ are indentical for starspots and ambient photosphere.  The composite mixture model for observed flux density can be written down as:

@@

The observed filling factor of starspots $c$ is simply the ratio:

@@

The starspot model includes two more free parameters than the standard \iancze\ model, namely $\teffb$ and $c$.  The addition of these two parameters makes the MCMC sampling much more correlated than it was before because the relative contribution of the starspot and ambient photosphere are nearly degnerate over small changes in effective temperature.  So the second important change to the spectral inference framework involves technical aspects of switching from sampling the nuisance and stellar parameters separately in a blocked Gibbs framework to ensemble sampling using \texttt{emcee} \citep{foreman13}.  The affine-invariant \texttt{emcee} ensemble sampler is more resilient to correlations among stellar temperatures and fill factor than the Metropolis-Hastings sampler used in Gibbs sampling.  The practical effect of this switch is that we obtain unique stellar parameter estimates $\vT_{o}$ for each spectral order $o$, whereas the \iancze\ strategy had the power to provide a single set of stellar parameters $\vT$ that was based on all $N_{\rm ord}$ spectral orders.  We combine the $N_{\rm ord}$ sets of inferences on $\vT_{o}$ with heuristics like taking the weighted average of point estimates of parameter distributions of the most reliable subsets of orders.  These lossy heuristics should regress to the mean with a large number of spectral orders, assuming each order delivers unbiased estimates of stellar parameters.

Further details of the spectral inference methodology are described in Appendix \ref{methods-details}.

@@

\subsection{IGRINS Spectroscopy}\label{sec:igrins} 
We acquired observations with the Immersion Grating Infrared Spectrograph, IGRINS \citep{gully12,park14} on the Harlan J. Smith Telescope at McDonald Observatory on 2015-11-18 $09^h$ UTC.  IGRINS is a high resolution near-infrared echelle spectrograph providing simultaneous $R\simeq45,000$ spectra over 1.48-2.48\um.  The spectrograph has two arms with 28 orders in $H-$band and 25 orders in $K-$band.  The data were reduced with the Pipeline Package\footnote{\url{https://github.com/igrins/plp}}.  Telluric correction was performed by dividing the spectrum of \name by a spectrum of HR 1237, an A0V star observed immediately before \name, both at airmass 1.1.  The broad hydrogen lines in the A0V star produced broad flux excesses in the spectrum of \name.  No effort was made to remove these flux excesses, which affect several spectral orders.

\subsection{ESPaDOnS Spectroscopy}
We used ESPaDOnS on CFHT to obtain twelve high resolution optical spectra of \name from 8-21 Jan.~2014 as part of the MATYSSE Large Program.  These spectra cover 3900$-$10000 \AA\ at $R\sim68,000$ and were obtained in spectropolarimetry mode.  The Zeeman Doppler Imaging obtained from these observations were analyzed by \citet{donati14}.  In this paper we concentrate on the intensity spectrum.  We apportioned the spectrum into a subsets of $N=35$ chunks, corresponding approximately to spectral order boundaries.
\todo{Flux calibration?}

@@

We performed spectral fitting on the ESPaDoNs spectrum acquired on 2014 January 11, the fourth spectrum from Table \ref{tbl_estimated_V}.  Full-spectrum fitting was performed separately on each of 35 spectral orders following the Metropolis-Hasting MCMC sampling procedure described in \iancze.  The spectral emulator was trained on stellar parameters in the range $\logg \in [3.5, 4.0]$, $\Z \in [-0.5, 0.5]$, and $\teff \in [3500, 4200]$.

The Metropolis-Hastings step sizes were tuned with several iterations of burn-in procedure and the final chains were visually checked for convergence.  Nine of the 35 available spectral orders failed to converge, possibly due to numerical artifacts arising from poor model fits.  For the other 26 spectral orders, the number of samples was generally much longer than the estimated integrated autocorrelation length.  Overlays of models with observed spectra show modest agreement, with the exception of conspicuous spectral line outliers.  No attempt was made to downweight spectral-line outliers as described in \iancze.

We computed the median value and 5$^{th}$ and $95^{th}$ percentiles of burned-in subsets of the MCMC samples described above.  Overall these 26 sets of point estimates for $\vT_{m}$, $\vP_m$ show relatively good agreement, with exceptions.  We have little constraint on $\logg$ and $\Z$ since full-spectrum fitting tends to wash out the weak signal contained in line wings for spectra with a finite level of random noise and systematic error.  Meanwhile, strong metal lines can bias estimates of $\Z$ at the expense of estimates for $\teff$.  We get consistent estimates for $v_z \sim 15.6 \pm 0.6$ km/s and $\vsini \sim 28.0 \pm 0.9$ km/s, with the exception of two spectral orders at $6920 <\lambda \;(\mathrm{nm})< 7057$ and $8473 < \lambda \;(\mathrm{nm}) < 8706$.  These two outlying spectral orders converged to unreasonably large values for $\vsini$ and $a_{\rm G}$, the amplitude of the Gaussian process covariance kernel.  One of these orders has an unusually small value for effective temperature, $\teff\sim3500$, comparable to the lower limit of the employed model grid.  Visual comparison of the observed spectrum for a range of fits for these orders shows that no single emulated spectrum could reasonably fit their spectrum because there are numerous large spectral outliers.  The more-well-behaved spectral orders show effective temperature point estimates in the range $\teff=4000\pm130$ K.  Figure \ref{fig:SingleTeffvsOrder} displays the $\teff$ point estimates with 5$^{th}$ and $95^{th}$ percentile error bars placed at the central wavelength of each of the 26 spectral orders.

@@

The effective temperature derived at different wavelength intervals exhibit a $\Delta \teff \sim 800$ K spread.  No single temperature can describe all the spectral lines present in the high resolution optical and IR spectra.  This discrepancy is circumstantial evidence for the detection of spectral lines attributable to starspots.  This circumstantial evidence should not be surprising given the previous evidence that \name possesses a large coverage fraction of starspots \citep{grankin08,donati14}.

%We performed a further plausibility check to constrain the effect size of starspots.  We generated flux calibrated spectra at two temperatures, and forward-modeled them to resemble the LkCa4 IGRINS and ESPaDOnS spectra in all other ways (\emph{i.e.} $\logg, \Z, \vt_{\rm ext}$), including noise and calibration parameters.  We coadded the spectra in a mixture model:  $ \mathsf{M}_{mix} = c \cdot \mathsf{M}_A(T_\textrm{eff,A}) + (1-c) \cdot \mathsf{M}_B(T_\textrm{eff,B})$.  We chose a fill factor of starspots of 30\%.  We then re-ran our single-temperature fitting procedure on the two-temperature, synthetic, noised-up data to see what stellar parameters our full-spectrum fitting procedure would na\"{\i}vely derive.

@@

Some care should be taken when directly comparing results between the ESPaDoNs and IGRINS spectra since they were not taken at the same time.  These spectra have comparable albeit different $\hat V$ of 12.90 and 12.83 for the ESPaDoNs and IGRINS spectra respectively.  Assuming the areal coverage fraction of spots was the same or greater during the ESPaDoNs spectrum acquisition, the optical bands and short-wavelength end of $H$-band are relatively insensitive to the spectral signatures of starspots in this system.  The long wavelength portion of $H-$band and all of $K-$band are more sensitive to starspot spectral signatures than the shorter wavelength portions.

The heightened sensitivity to starspot spectral lines as a function of wavelength can be understood in the following way.  The effective temperature of a starspot is, by definition, cooler than its surrounding photosphere, and will therefore have a longer wavelength of peak emission, according to Wein's displacement law.  So the ratio of flux density between starspot and bulk photosphere as a function of wavelength will generally increase with wavelength until it assymptotes to a fixed value in the Rayleigh-Jeans tail \citep{wolk96}.

%In Figure XX we plot the ratio of black body flux density at temperatures 4100 K and 3300 K with a red line.  The blue line is the ratio of two smoothed \PHOENIX model spectra with 4100 K and 3300 K, solar metalicity, and $\logg=3.5$.  The flux density ratio is the flux per solid angle, so the delivered flux ratio for a finite sized starspot will depend on the (flux-weighted) solid angle of the starspot relative to the (flux-weighted) solid angle of the stellar disk, vis-a-vis Equations \ref{eqn:Mix_M} and \ref{eqn:fill_factor}.  
%\todo{Put in the flux density ratio plot.}
